---
title: "traitstrap-workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{traitstrap-worklfow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(traitstrap)
library(ggplot2)
```


## The aim of traitstrap
Traitstrap is an R package that estimates the moments of community trait distributions using a bootstrapping approach.
The package has three main functions: (i) a trait imputation function which uses a hierarchical sampling design to impute missing or incomplete trait data and can take into account experimental sampling designs.
(ii) a resample method using bootstrapping (parametric or non-parametric method)  to calculate community weighted trait values and the statistical moments (mean, variance, skewness and kurtosis).
(iii) a summary function that summarizes the trait moments and calculates confidence intervals.


## Organize your data

To use traitstrap you need to have two datasets: 
* one dataset with information on the abundance/biomass/size of your community, which is used to weight species traits by dominance.
* one dataset with the traits for each species (or as many species and individuals you have data for) in your community.

The datasets need to be organized in a **tidy** and **long** format.
It needs to contain certain columns, but the naming of the columns are up to the user.

Let us have a look the two datasets in this example.

The **community data** should have information about the species and their abundance or size in the community.
This dataset will be used to weigh the traits by species dominance.

In this example the data has a column with the species names (e.g. **Taxon**), a column with the cover of each species per plot (e.g. **Cover**) and two columns with information about the **hierarchy** (i.e. **Site** and **PlotID**).


```{r comm-data, echo=FALSE, eval=TRUE}
community
```

The **trait data** contains information about the traits and the trait values for each species (and individual) in the commuinty.
The data should be organized in the same way as the community data should have corresponding columns.
In this example the trait data has the following columns: **Taxon**, **Site** and **PlotID** as well as columns with information about the traits and their values **Trait**, **Value**.

```{r trait-data, echo=FALSE, eval=TRUE}
trait
```


## Trait imputation

The **trait_impute function** uses a hierarchical sampling design, which allows it to account for incomplete trait collections, traits from different spatial or temporal levels (i.e. local traits vs. databases) and/or experimental designs. 

The first two mandatory arguments in the function are the two datasets:
**comm** and **traits**

Next you need to define four columns in your datasets:
* **abundance** which is a measure of dominance of your species in your community dataset. This can be abundance, cover, biomass, size, etc.
* **taxon_col** is the column in your community and trait data that define the species names.
* **trait_col** is the column in your trait data that defines the traits
* **value_col** is the column in your trait data that defines the trait values

All the other arguments in traitstrap are not mandatory.

With **scale_hierarchy** you can define the levels at which the traits have been collected and their order starting with the highest level (e.g. global database, region, site, block, plot).
In the example below we have **Site** and **PlotID**.

The trait_impute function will choose if available a trait value from the lowest level, i.e. species X from plot A and if no trait is available from that level, it will move up the hierarchy and choose a trait from species X from plot B at the same site.
If there is no trait available from species X in the same site, it will choose a trait value from another site.

The argument **min_n_in_samples** allows users to define the minimum number in sample at each level for the trait imputation.
If the minimum number is not reached, trait values from the next level will also be imputed, to avoid sampling the same individual several times, which will result in unrealistic variances.
The default value is 5

In the **other_col** arumgent you can define columns in the trait dataset that are not important for the trait imputation but that you want to keep.


```{r trait-impute, echo=TRUE, eval=TRUE}

trait_imputation <- trait_impute(
    # input data
    comm = community,
    traits = trait,
    
    # specifies mandatory columns in your data
    abundance_col = "Cover",
    taxon_col = "Taxon",
    trait_col = "Trait",
    value_col = "Value",
    
    # specifies hierarchy
    scale_hierarchy = c("Site", "PlotID"),
    
    # min number of samples
    min_n_in_sample = 3
  )

trait_imputation
```


Traitstrap also allows to include taxonomy and experimental design in the trait imputation step.

With **taxon_col** you can define a hierarchy for the taxonomy.
If traits for a specific species are not available, traits from the same genus will be imputed.
For this a list of the taxonomic hierarchy has to be defined (e.g. "Taxon", "Genus", "Family" etc.).

The argument **treatment_col** allows to incorporate an experimental design where traits are imputed from the same experimental treatment or the first factor level, which is assumed to be the control.
Therefore, it is important to order the levels of a treatment in the right order, i.e. the first level has to be the control.
The imputation step can be defined at certain level using the **treatment_level** argument.
Depending on the experimental design it might make sense to impute traits at a certain level, e.g. block or site.

Here is an example how to include taxonomy and experimental desing in the trait imputation function.

```{r trait-impute2, echo=TRUE, eval=FALSE}

trait_imputation2 <- trait_impute(
    comm = community,
    traits = trait,
    
    abundance_col = "Cover",
    
    # defining taxonomic hierarchy
    taxon_col = c("Taxon", "Genus"),
    
    trait_col = "Trait",
    value_col = "Value",
    
    scale_hierarchy = c("Site", "PlotID"),
    min_n_in_sample = 3
    
    # specifying experimental design
    treatment_col = "Treatment",
    treatment_level = "Site",

  )
```




## Non-parametric bootstrapping

The output of the trait imputation function can be used to do a **non-parametric bootstrapping** using the **trait_np_bootstrap** function.
This is a resampling method to calculate community weighted trait values.
The observed traits are re-sampled in proportion to their weights (e.g. by the abundance of the species).
Values across all individuals in a community are resampled n times (**sample_size**; the default is 200) to incorporate the full spectrum of trait variation, generating n number (**nrep**; the default is 100) of trait distributions. 
From these distributions the statistical moments are estimated: **mean**, **variance**, **skewness** and **kurtosis**.


```{r non-parap-boot, echo=TRUE, eval=TRUE}

# run non-parametric bootstrapping
np_bootstrapped_moments <- trait_np_bootstrap(
  trait_imputation, 
  nrep = 200
  )

np_bootstrapped_moments
```


Finally, the statistical moments can be summarized and the confidence intervalls calcualted using the **trait_summarise_boot_moments** function.
The input variable is the output from the non-parametric boostrapping function.

<!-- This needs checking and maybe some more details on how to get 99% CI etc. -->
The **sd_mult** and **ci** argument can be used to define the desired confidence intervall.
The default is `sd_mult = 1` which gives the 95% confidence interval.

```{r summarize, echo=TRUE, eval=TRUE}

# summarizes bootstrapping output
sum_boot_moment <- trait_summarise_boot_moments(
  np_bootstrapped_moments
  )

sum_boot_moment
```


## Parametric bootstrapping

Traitstrap also has the option to run a **parametric bootstrapping**.

The **trait_fit_distributions** function fits parametric distributions for each species-by-trait combination at the finest scale of the user-supplied hierarchy. 
This function takes as input:
1) an object of class imputed traits (as produced by the function trait_impute), and 
2) the type of distribution to be fitted.
Either a single distribution type can be used for all traits, or traits can be assigned specific distributions types by supplying the function with a named list of traits.
Currently supported distribution types are normal, log-normal, and beta.
The function returns a dataframe containing fitted distribution parameters.

```{r fit-dist, echo=TRUE, eval=TRUE}

# fit distributions
fitted_distributions <- trait_fit_distributions(
  imputed_traits = trait_imputation,
  distribution_type = "lognormal"
  )

fitted_distributions
```


The **trait_parametric_bootstrap** function is a parametric analog of the trait_np_bootstrap function.
It takes in fitted trait distributions produced by **trait_fit_distributions** and randomly samples from among the fitted distributions proportionally to species abundances.
As with trait_np_bootstrap, the number of samples per replicated draw are specified with the parameter sample_size, and the number of replicated draws is specified by the parameter **nrep**.
The output of trait_parametric_bootstrap can be summarized using **trait_summarize_boot_moments** (see above).

```{r para-boot, echo=TRUE, eval=TRUE}

# run paramteric bootstrapping
p_bootstrapped_moments <- trait_parametric_bootstrap(
    fitted_distributions = fitted_distributions, 
    nrep = 200
    )

p_bootstrapped_moments
```


<!-- Could also show code how to plot the raw distributions -->


## Check your data

Traitstrap has a couple of functions to check your data.

The **coverage_plot** function shows the trait coverage of the community for each level.
Basically, this function summarizes from which level the traits are imputed, and how much coverage of the community is reached.
<!-- Add an explanation about why this plot is important! -->
This can be important plot to show, because we know that the coverage should be at least X% (ref to the paper!)

```{r coverage-plot, echo=TRUE, eval=TRUE}

# show coverage plot
autoplot(trait_imputation)

```


Another important information is to know of which taxa traits are missing.
This can be useful if the data sampling is not finished and you want to know which species should be sampled.
The function also tells you the maximal abundance of each missing species, and gives you useful information if the missing species are abundant or rare.

Traitstrap has a function **trait_missing** which gives you a table with all missing values.


```{r missing-traits, echo=TRUE, eval=TRUE}

# list missing traits
trait_missing(imputed_trait = trait_imputation,
              comm = community)

```

